|

### 2023-04-06 manuscript/paper.Rmd
Added some intro and added a fun title.

---
title: |
  |  \Large Eras in baseball: Change point analysis fun title. 
author: | 
  | \large Mena Whalen \vspace{-1.1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  | \normalsize [`mwhalen3@luc.edu`](mailto:mwhalen3@luc.edu) \vspace{-1mm}
  |
  | \large Gregory J. Matthews \vspace{-1.1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  | \normalsize [`gmatthews1@luc.edu`](mailto:gmatthews1@luc.edu) \vspace{-1mm}
abstract: |
  | Baseball is some weird and wild shit. \vspace{2mm}
  | *Keywords*: change point analysis, baseball, 
bibliography: references.bib
fontsize: 12pt
link-citations: true
linkcolor: cyan
urlcolor: cyan
output:
  pdf_document:
    df_print: kable
    number_sections: true
    keep_tex: true
header-includes:
 \usepackage{setspace}
 \setstretch{1.15}
 \usepackage{float}
 \floatplacement{figure}{t}
---

@@

@@

Baseball is not the only sport where this type of analysis has been applied.  @PalaciosHuerta2004 looked for structural changes in soccer using data from British soccer leagues through 1996. @FortLee2007 looked for structural breaks in major North American sports other than baseballl; that is basketball (NBA), hockey (NHL), and American football (NFL).  WHAT DID THEY FIND?

All of the previous work mentioned here focuses on change point analysis in ***univariate*** time series. However, recent methodolgical developments in change point analysis allow for change point analysis in ***multivariate*** time series, which is the focus of this current manuscript.  This work leverages techniques such as the Double CUSUM Binary Segmentation algorithm (@Cho2016) and the Sparsified Binary Segmentation algorithm (@ChoFryzlewwicz2014) to look for change points in Major League Baseball at two main levels.  First, we look for structural changes in the league as a whole across teams to empirically define different eras in baseball.  Second, we look for change points within a team to determine eras within a team.  In essence, this second goal can be used to identify the beginning and end of so called "dynasties", periods of sustained excellent performance by a team.  In addition, we can also identify the opposite, sustained periods of poor performance which we will term ERAS of SUCKITUDE.

Just Notes: 
Palacios-Huerta I. (2004). Structural changes during a Century of the World’s most popular sport. Statistical Methods and Applications, 13, 241–258
@PalaciosHuerta2004 Structural Breaks in soccer

In contrast to their methods, we use  more modern methods Double CUSUM Binary Segmentation algorithm (@Cho2016) and the Sparsified Binary Segmentation algorithm (@ChoFryzlewwicz2014).

https://onlinelibrary.wiley.com/doi/epdf/10.1111/j.1465-7295.2007.00026.x
Fort and Lee 2007: Looks for structural breaks in NBA, NHL, and NFL

https://www.baseball-reference.com/bullpen/History_of_baseball_in_the_United_States

https://thesportjournal.org/article/examining-perceptions-of-baseballs-eras/

@@

https://pubmed.ncbi.nlm.nih.gov/30747582/?

The goal of this work is to use multivariate change point analysis to identify change points in MLB as a whole to objectively identify eras in the games history, but also to look for change points to identify team specific eras.

The work that is the most closely related to our own is that of @Groothius2017, which looked for structural breaks (Note: the terms "structural breaks" and "change points" are used interchangeably in this manuscript) in univariate times series.

univariate time series of the means and SD's of several statistics (i.e. SLUG, HR rate, RBI rate, and batting average) to look for structural breaks via the Lagrange Multiplier (LM) unit root test @LeeandStrazicich2003.  Notably they find structural breaks in slugging percentage in 1921 and 1992.  In contrast to their methods, we use  more modern methods Double CUSUM Binary Segmentation algorithm (@Cho2016) and the Sparsified Binary Segmentation algorithm (@ChoFryzlewwicz2014).

@Woltring2018
From Woltring:
"Baseball has endured much change over the course of its history, and because of constant change, the modern era of baseball has been segmented into six distinct sub-eras. A common list presented at Baseball-Reference described the eras as the Dead Ball Era (1901-1919), the Live Ball Era (1920-1941), the Integration Era (1942-1960), the Expansion Era (1961-1976), the Free Agency Era (1977-1993) and the Long Ball/Steroid Era (1994-2005) (17). This study runs through the 2011 season and a seventh era will be added and labeled the Post Steroid Era (2006-2011)"

>>>>

---
title: |
  |  \Large Eras in baseball: Change-Ups and Change Points: An Exploration of Baseball's Historic Eras
author: | 
  | \large Mena Whalen \vspace{-1.1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  | \normalsize [`mwhalen3@luc.edu`](mailto:mwhalen3@luc.edu) \vspace{-1mm}
  |
  | \large Gregory J. Matthews \vspace{-1.1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  | \normalsize [`gmatthews1@luc.edu`](mailto:gmatthews1@luc.edu) \vspace{-1mm}
abstract: |
  | Baseball is some weird and wild shit. \vspace{2mm}
  | *Keywords*: change point analysis, baseball, 
bibliography: references.bib
fontsize: 12pt
link-citations: true
linkcolor: cyan
urlcolor: cyan
output:
  pdf_document:
    df_print: kable
    number_sections: true
    keep_tex: true
header-includes:
 \usepackage{setspace}
 \setstretch{1.15}
 \usepackage{float}
 \floatplacement{figure}{t}
---

@@

More specifically, @Woltring2018 mentions six eras of modern baseball": 
"Baseball has endured much change over the course of its history, and because of constant change, the modern era of baseball has been segmented into six distinct sub-eras. A common list presented at Baseball-Reference described the eras as the Dead Ball Era (1901-1919), the Live Ball Era (1920-1941), the Integration Era (1942-1960), the Expansion Era (1961-1976), the Free Agency Era (1977-1993) and the Long Ball/Steroid Era (1994-2005) (17)."  @Woltring2018 notes that they name a seventh era after 2006, which they term the Post Steroid Era.

@@

Baseball is not the only sport where this type of analysis has been applied.  @PalaciosHuerta2004 looked for structural changes in soccer using data from British soccer leagues through 1996.  They had data from division I (Premier League) and II starting in the late 1800s, and data from lower divisions III and IV from just after WWII starting in 1947.  They identify a number of change points.  Notably they identify a change point in the mean of margin of victory in 1925 related to the change in the definition of offsides (changed from 3 players to 2 players), an change points in the variability of number of goals in the early 1980s and 1992 corresponding to the change in number of points for a win (i.e. went from 2 points to 3 points) and a change in the backpass rule, respectively.  @FortLee2007 looked for structural breaks in major North American sports other than baseball (i.e. basketball (NBA), hockey (NHL), and American football (NFL)).  They identified a number of change points related to competitive balance in each sport that often, though not always, correspond to league expansion, league mergers, or other major events in a sport (e.g. increased number of foreign players in the NBA in the late 1990s/early 2000s).

<!-- WHAT DID THEY FIND?   ADD MORE HERE.  -->

All of the previous work mentioned here focuses on change point analysis in ***univariate*** time series. However, recent methodolgical developments in change point analysis allow for change point analysis in ***multivariate*** time series, which is the focus of this current manuscript.  This work leverages techniques such as the Double CUSUM Binary Segmentation algorithm (@Cho2016) and the Sparsified Binary Segmentation algorithm (@ChoFryzlewwicz2014) to look for change points in Major League Baseball at two main levels.  First, we look for structural changes in the league as a whole across teams to empirically define different eras in baseball.  Second, we look for change points within a team to determine eras of a team.  This second analysis can be used to identify the beginning and end of so called "dynasties", periods of sustained excellent performance by a team.  In addition, we can also identify the opposite, sustained periods of poor performance.

<!-- Just Notes:  -->
<!-- Palacios-Huerta I. (2004). Structural changes during a Century of the World’s most popular sport. Statistical Methods and Applications, 13, 241–258 -->
<!-- @PalaciosHuerta2004 Structural Breaks in soccer -->

<!-- @Nieswiadomy2012 Was there a structural break in Barry Bonds’s Bat?  Looking for structural breaks in Bonds OPS times series. Use the uni root test on monthly data from 86-07.  The identify break points in June 1993 and Sept 2000.  -->

<!-- In contrast to their methods, we use  more modern methods Double CUSUM Binary Segmentation algorithm (@Cho2016) and the Sparsified Binary Segmentation algorithm (@ChoFryzlewwicz2014). -->

<!-- https://onlinelibrary.wiley.com/doi/epdf/10.1111/j.1465-7295.2007.00026.x -->
<!-- Fort and Lee 2007: Looks for structural breaks in NBA, NHL, and NFL -->

<!-- https://www.baseball-reference.com/bullpen/History_of_baseball_in_the_United_States -->

<!-- https://thesportjournal.org/article/examining-perceptions-of-baseballs-eras/ -->

@@

<!-- https://pubmed.ncbi.nlm.nih.gov/30747582/? -->

<!-- The goal of this work is to use multivariate change point analysis to identify change points in MLB as a whole to objectively identify eras in the games history, but also to look for change points to identify team specific eras.   -->

<!-- The work that is the most closely related to our own is that of @Groothius2017, which looked for structural breaks (Note: the terms "structural breaks" and "change points" are used interchangeably in this manuscript) in univariate times series.   -->

<!-- univariate time series of the means and SD's of several statistics (i.e. SLUG, HR rate, RBI rate, and batting average) to look for structural breaks via the Lagrange Multiplier (LM) unit root test @LeeandStrazicich2003.  Notably they find structural breaks in slugging percentage in 1921 and 1992.  In contrast to their methods, we use  more modern methods Double CUSUM Binary Segmentation algorithm (@Cho2016) and the Sparsified Binary Segmentation algorithm (@ChoFryzlewwicz2014). -->
